---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
---

```{r opts, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "figures/"
)
```

## Loading and preprocessing the data

First load libraries
```{r libraries, warning=FALSE}
library(scales)
library(ggplot2)
library(knitr)
```

Then loading data and converting date , with data summary
```{r load data}
unzip("./activity.zip")
activities <- read.csv("./activity.csv")

activities$date <- as.Date (activities$date,"%Y-%m-%d")
summary(activities)

```



## What is mean total number of steps taken per day?

* Calculate the total number of steps taken per day
```{r calculate_total_step_per_day}
daysteps <- setNames(aggregate(activities$steps,by=list(activities$date),sum,na.rm=T),c("date","steps"))
```

*  Make a histogram of the total number of steps taken each day  
```{r histogram_steps_by_day}
gg <- ggplot(data=daysteps, aes(steps)) + geom_histogram(breaks=seq(0, 25000, by=5000),alpha=.5,col="white") + labs(title = "Histogram - Total daily number of steps")
print(gg)
                                                              
```

**Barplot**
```{r barplot_steps_by_day}
gg <- ggplot(data = daysteps, aes(x=date,y=steps)) + geom_bar(stat = 'identity') + scale_x_date(date_breaks = '1 week',date_labels = "%b %d") + labs(title = "Barplot - Total daily number of steps")
print(gg)
                                                              
```


* Calculate and report the mean and median of the total number of steps taken per day
```{r mean and median steps per day}
mea <- mean(daysteps$steps)
med <- median( daysteps$steps)

```
The mean and median of the total number of steps taken per day are respectively **`r format(mea)`** and **`r med`**.

## What is the average daily activity pattern?

```{r plot_average_daily_activitiy_pattern}
dayintervals <- aggregate(steps ~ interval, mean, data=activities, na.rm=T)

gg <- ggplot(data = dayintervals, aes(x=interval,y=steps)) + geom_line() + labs(title = "average daily activity pattern - steps per interval")
print(gg)

```


```{r}
m <- subset(dayintervals,steps==max(dayintervals$steps))
```
The 5-minute interval which, on average across all the days in the dataset, contains the maximum number of steps is the number **`r m[1]`**, with an average of **`r format(m[2])`** steps.

## Imputing missing values
Since there are a number of days/intervals where there are missing values (coded as NA), they  may introduce bias into some calculations or summaries of the data.

* Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
totNA <- sum(is.na(activities$steps))
```
The total number of observations with missing values is **`r totNA `**.

* Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r filling missing value}
activitiesNoNA <- read.csv("./activity.csv")
activitiesNoNA$date <- as.Date (activitiesNoNA$date,"%Y-%m-%d")

# Devising a strategy for filling in all of the missing values in the dataset,
# i used the mean for that 5-minute interval.
for(i in seq_along(activitiesNoNA$steps)){
  if(is.na(activitiesNoNA$steps[i])){
    activitiesNoNA$steps[i] <- subset(dayintervals,interval==activitiesNoNA$interval[i])[,2]
  }
}
```

* Make a histogram of the total number of steps taken each day.  
```{r histogram_steps_per_day_with_imputed_values}
daystepsNoNA <- aggregate(steps ~ date, sum, data=activitiesNoNA, na.rm=T)
gg <- ggplot(data=daystepsNoNA, aes(steps)) + geom_histogram(breaks=seq(0, 25000, by=5000),alpha=.5,col="white") + labs(title = "Histogram - Total daily number of steps ")
print(gg)
                                                              
```

```{r barplot_steps_per_day_with_imputed_values}
gg <- ggplot(data = daystepsNoNA, aes(x=date,y=steps)) + geom_bar(stat = 'identity') + scale_x_date(date_breaks = '1 week',date_labels = "%b %d") + labs(title = "Barplot - Total daily number of steps - with imputed values")
print(gg)
```

* Calculate and report the mean and median total number of steps taken per day
```{r}
meaNoNA <- mean(daystepsNoNA$steps)
medNoNA <- median( daystepsNoNA$steps)

```
The mean and median of the total number of steps taken per day after imputed missing values are respectively **`r format(meaNoNA)`** and **`r format(medNoNA)`**.

* Do these values differ from the estimates from the first part of the assignment?  
The values differ from the first part of assignment respectively: mean by **`r format(meaNoNA-mea)`** and median by **`r format(medNoNA-med)`**.


* What is the impact of imputing missing data on the estimates of the total daily number of steps?
Days impacted by imputing missing data  
```{r table, results="asis" }
df <-setNames(cbind(daysteps,daystepsNoNA$steps),c("date","stepswNA","stepswoNA")) 

kable(subset(df,(stepswoNA-stepswNA)>0), format = 'markdown',caption="Days impacted by imputing missing data",col.names = c("date","total steps with NA","total steps after imputing missing data")) 
```

## Are there differences in activity patterns between weekdays and weekends?

Use the dataset with the filled-in missing values for this part.

* Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r factor daytype}
activitiesNoNA$daytype <- factor(ifelse(as.POSIXlt(activitiesNoNA$date)$wday %in% c(0,6),"weekend","weekday"))
```

* Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r panel_plot_on_daytype}
datedf <- aggregate(steps ~  interval + daytype , FUN = mean, data=activitiesNoNA)
gg <- ggplot(data=datedf, aes(x=interval,y=steps))+geom_line()+ facet_grid(daytype~.)
print(gg)

```
